---
title: "Coursera Practical Machine Learning Class - Project"
author: "Loïc BERTHOU"
date: "August 23, 2015"
output: html_document
---

```{r, echo=FALSE}
set.seed(20150823)
```

## Executive Summary

This document aims at demonstrating how we can use devices such as Jawbone Up, Nike FuelBand, and Fitbit in order to track the quality of exercises instead of quantity like it is the case most of the time. In our specific case, we are trying to predict if an athlete has perform barbell lifts correctly or incorrectly.

## Data 

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

## Data Cleaning

We will start by loading the _training_ and _testing_ datasets.

```{r, cache=TRUE, cache=TRUE}
trainingSet = read.csv("pml-training.csv", header = TRUE, stringsAsFactors = FALSE, na.strings = c("", "#DIV/0!", "NA"))
testingSet = read.csv("pml-testing.csv", header = TRUE, stringsAsFactors = FALSE, na.strings = c("", "#DIV/0!", "NA"))
```

Let's have a brief overview of them.

```{r}
str(trainingSet)
str(testingSet)
```

Our initial exploratory data analysis shows that some columns contain N/A values. Let's first count these N/A values.

```{r, cache=TRUE}
isNaMatrixTraining <- apply(trainingSet, 1, is.na)
countNaMatrixTraining <- apply(isNaMatrixTraining, 1, sum)
countNaMatrixTraining <- as.data.frame(countNaMatrixTraining)
countNaMatrixTraining

isNaMatrixTesting <- apply(testingSet, 1, is.na)
countNaMatrixTesting <- apply(isNaMatrixTesting, 1, sum)
countNaMatrixTesting <- as.data.frame(countNaMatrixTesting)
countNaMatrixTesting
```

Since the number of N/A values is non-neglectable, we're going to remove all these columns.

```{r}
# Combine the columns that contain N/A values.
noNaMatrix <- as.data.frame(cbind(countNaMatrixTraining[,1] == 0, countNaMatrixTesting[,1] == 0))
names(noNaMatrix) <- c("training", "testing")
noNaMatrix$combined <- noNaMatrix$training & noNaMatrix$testing

trainingSetClean <- trainingSet[, noNaMatrix$combined]
testingSetClean <- testingSet[, noNaMatrix$combined]
```

We will also remove the timestamp columns that will not be used in the analysis.

```{r}
trainingSetClean <- subset(trainingSetClean, select=-c(X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, new_window, num_window))
testingSetClean <- subset(testingSetClean, select=-c(X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, new_window, num_window))
```

Now we will make the factor variables as factor columns.

```{r}
trainingSetClean$classe <- as.factor(trainingSetClean$classe)
```

## Data Modeling

We first load the libraries for machine learning and parallelization (to set-up our system to use multiple cores).

```{r, message=FALSE}
library(caret)
library(doParallel)
library(randomForest)

# Create the cluster of workers
cl <- makeCluster(detectCores())
registerDoParallel(cl)
```

### Sub-Sampling

Since we know that our training set contains many observations (`r nrow(trainingSetClean)` rows), we fear that the training algorithm would take a long time. We will then first use a small sample of the original training sample to investigate the best model to use in our machine learning algorithm.

```{r}
# Sub-sampling 2000 rows randomly
sample1 <- trainingSetClean[sample(1:nrow(trainingSetClean), 2000, replace=FALSE),]

# Creating the training and testing data partitions (75% training)
dataPartitions <- createDataPartition(sample1$classe, p=0.75, list=FALSE)
training1 <- sample1[dataPartitions[, 1], ]
testing1 <- sample1[-dataPartitions[, 1], ]

# Checking the size our new sub datasets
dim(training1)
dim(testing1)
```

### The model is being generated with the 'Random Forest' method.

```{r, cache=TRUE}
timeRf.start <- Sys.time()
#fitRfControl <- trainControl(method = "cv", number = 5)
#fitRf <- train(classe ~ ., data = training1, method = "rf", trControl = fitRfControl, prox = TRUE)
fitRf <- randomForest(classe ~ ., data = training1)
timeRf.end <- Sys.time()
timeRf.taken <- timeRf.end - timeRf.start
fitRf
timeRf.taken
```

Let's predict the values on the sub-testing sample anc compare them to the reference.

```{r}
predictRf <- predict(fitRf, testing1)
confRf <- confusionMatrix(predictRf, testing1$classe)
confRf
```

With this small sample, the model was generated quite fast (`r sprintf("%.1f", timeRf.taken)` seconds) and we already have an accuracy rate of `r sprintf("%.1f", confRf$overall[1] * 100)`% with a 95% confidence interval between `r sprintf("%.1f", confRf$overall[3] * 100)`% and `r sprintf("%.1f", confRf$overall[4] * 100)`% !
This one looks like our winner but let's still investigate other methods.

### The model is being generated with the Caret package and the 'Boosted Tree' method.

```{r, cache=TRUE}
timeGbm.start <- Sys.time()
fitGmbControl <- trainControl(method = "repeatedcv", number = 5, repeats = 5)
fitGbm <- train(classe ~ ., data = training1, method = "gbm", trControl = fitGmbControl, verbose = FALSE)
timeGbm.end <- Sys.time()
timeGbm.taken <- timeGbm.end - timeGbm.start
fitGbm
timeGbm.taken
```

Let's predict the values on the sub-testing sample and compare them to the reference.

```{r}
predictGbm <- predict(fitGbm, testing1)
confGbm <- confusionMatrix(predictGbm, testing1$classe)
confGbm
```

This method is slower (`r sprintf("%.1f", timeGbm.taken)` seconds) and the accuracy rate is below the previous method at `r sprintf("%.1f", confGbm$overall[1] * 100)`% with a 95% confidence interval between `r sprintf("%.1f", confGbm$overall[3] * 100)`% and `r sprintf("%.1f", confGbm$overall[4] * 100)`%.

### The model is being generated with the Caret package and the 'Prediction Tree' method.

```{r, cache=TRUE}
timeRpart.start <- Sys.time()
fitRpart <- train(classe ~ ., method="rpart", data=training1)
timeRpart.end <- Sys.time()
timeRpart.taken <- timeRpart.end - timeRpart.start
print(fitRpart$finalModel)
timeRpart.taken
```

Let's predict the values on the sub-testing sample anc compare them to the reference.

```{r}
predictRpart <- predict(fitRpart, testing1)
confRpart <- confusionMatrix(predictRpart, testing1$classe)
confRpart
```

This method is fast (`r sprintf("%.1f", timeRpart.taken)` seconds) but the accuracy rate is bad at `r sprintf("%.1f", confRpart$overall[1] * 100)`% with a 95% confidence interval between `r sprintf("%.1f", confRpart$overall[3] * 100)`% and `r sprintf("%.1f", confRpart$overall[4] * 100)`%.

### Choosing the model

After trying out a few different models, the **Random Forest** seems like the most accurate. We will then run the training on the whole training set to create the final model used for our prediction.

We will first divide our origianl training set into two parts for the training and final validation. We are using 85% of the original training set for training and the 15% left will be used for cross validation.

```{r, cache=TRUE}
inTrain <- createDataPartition(trainingSetClean$classe, p = 0.85, list = FALSE)
trainingSetRf <- trainingSetClean[inTrain, ]
validationSetRf <- trainingSetClean[-inTrain, ]

time.start <- Sys.time()
# I am calling the randomForest function directly because the train() function makes it too slow
#fitControl <- trainControl(method = "cv", number = 5)
#fit <- train(classe ~ ., data = trainingSetRf, method = "rf", trControl = fitControl, ntree = 10)
fit <- randomForest(classe ~ ., data = trainingSetRf, ntree = 500)
time.end <- Sys.time()
time.taken <- time.end - time.start
fit
time.taken

# Good practice to shut down the workers.
stopCluster(cl)
```

Now we can predict the class for the validation dataset.

```{r}
validationSetRf$prediction <- predict(fit, validationSetRf)
confFit <- confusionMatrix(validationSetRf$prediction, validationSetRf$classe)
confFit
```

By validating against our validation set, we find that the model was accurate at `r sprintf("%.1f", confFit$overall[1] * 100)`% with a 95% that it will be accurate between `r sprintf("%.1f", confFit$overall[3] * 100)`% and `r sprintf("%.1f", confFit$overall[4] * 100)`%.

Now we can predict the class for the final test dataset.

```{r}
testingSetClean$classe <- predict(fit, testingSetClean)
```

We will extract the result data that will be submitted for grading.

```{r}
results <- subset(testingSetClean, select = c(problem_id, classe))
results
```

And generate the result files.

```{r}
pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}
pml_write_files(results$classe)
```

## Note on optimisation of the methods

After having had problems of performance, I have discovered that using randomForest was better for performance and did not affect the results.

Some other improvements are possible. As you can see in the plot below, I could reduce the _ntree_ parameter to make the calculation faster and still have a very good accuracy (if ntree > 200)

```{r}
plot(fit)
```

