---
title: "Coursera Practical Machine Learning Class - Project"
author: "Loïc BERTHOU"
date: "August 23, 2015"
output: html_document
---

```{r, echo=FALSE}
set.seed(20150823)
```

## Executive Summary

This document aims at demonstrating how we can use devices such as Jawbone Up, Nike FuelBand, and Fitbit in order to track the quality of exercises instead of quantity like it is the case most of the time. In our specific case, we are trying to predict if an athlete has perform barbell lifts correctly or incorrectly.

## Data 

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

## Data Cleaning

We will start by loading the _training_ and _testing_ datasets.

```{r, cache=TRUE}
trainingSet = read.csv("pml-training.csv", header = TRUE, stringsAsFactors = FALSE)
testingSet = read.csv("pml-testing.csv", header = TRUE, stringsAsFactors = FALSE)
```

Let's have a brief overview of them.

```{r}
str(trainingSet)
str(testingSet)
```

Our initial exploratory data analysis shows that some columns contain N/A values. Let's first count these N/A values.

```{r, cache=TRUE}
isNaMatrixTraining <- apply(trainingSet, 2, is.na)
countNaMatrixTraining <- apply(isNaMatrixTraining, 2, sum)
countNaMatrixTraining <- as.data.frame(countNaMatrixTraining)
countNaMatrixTraining

isNaMatrixTesting <- apply(testingSet, 1, is.na)
countNaMatrixTesting <- apply(isNaMatrixTesting, 1, sum)
countNaMatrixTesting <- as.data.frame(countNaMatrixTesting)
countNaMatrixTesting
```

Since the number of N/A values is non-neglectable, we're going to remove all these columns.

```{r}
trainingSetClean <- trainingSet[, countNaMatrixTesting[,1] == 0]
testingSetClean <- testingSet[, countNaMatrixTesting[,1] == 0]
```

We will also remove the timestamp columns that will not be used in the analysis.

```{r}
trainingSetClean <- subset(trainingSetClean, select=-c(X, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, new_window, num_window))
testingSetClean <- subset(testingSetClean, select=-c(X, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, new_window, num_window))
```

Now we will make the factor variables as factor columns.

```{r}
trainingSetClean$user_name <- as.factor(trainingSetClean$user_name)
trainingSetClean$classe <- as.factor(trainingSetClean$classe)

testingSetClean$user_name <- as.factor(testingSetClean$user_name)
# testingSetClean$classe <- as.factor(testingSetClean$classe) # This column doesn't exist in testing set

```
